% eilmer3-theory-book.tex
% PJ, March 2007 -- elmer2 version
%     September 2008 -- eilmer3 version
%     May 2010 -- theory book adaption
%
\documentclass[12pt,a4paper,twoside]{article}
\usepackage[body={16cm,24.5cm}]{geometry}
\usepackage{times}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{pstricks}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{lscape}
\usepackage{longtable}
\lstset{basicstyle=\ttfamily\scriptsize,identifierstyle=,keywordstyle=}

%------------------------------------------------------------------
% a couple horizontal bars to delimit embedded code
% the width suits the page size set above and
% the mathmode eliminates spaces between the three elements
\newcommand{\topbar}{\ensuremath{
    \rule{0.1mm}{2.0mm} \rule[2.0mm]{159.5mm}{0.1mm} \rule{0.1mm}{2.0mm}
}}
\newcommand{\bottombar}{\ensuremath{
    \rule{0.1mm}{2.0mm} \rule{159.5mm}{0.1mm} \rule{0.1mm}{2.0mm}
}}
\newcommand{\topbarshort}{\ensuremath{
    \rule{0.1mm}{2.0mm} \rule[2.0mm]{149.5mm}{0.1mm} \rule{0.1mm}{2.0mm}
}}
\newcommand{\bottombarshort}{\ensuremath{
    \rule{0.1mm}{2.0mm} \rule{149.5mm}{0.1mm} \rule{0.1mm}{2.0mm}
}}
%------------------------------------------------------------------

\title{
    Eilmer's Theory Book: Basic Models for Gas Dynamics and Thermochemistry.
}
\author{
    Mechanical Engineering Report 2010/09\\
    P. A. Jacobs\thanks{Queensland Geothermal Energy Centre of Excellence, The University of Queensland, Brisbane, Australia.}, 
    R. J. Gollan\thanks{Centre for Hypersonics, The University of Queensland, Brisbane, Australia.},
    A. J. Denman, B. T. O'Flaherty, \\
    D. F. Potter, P. J. Petrie-Repar and I. A. Johnston.
}
% \date{May 2010}
\makeindex

\begin{document}
\maketitle

\begin{abstract}
\texttt{Eilmer3} is an integrated collection of programs for the simulation of transient
compressible flow in two and three spatial dimensions.
It is based on a finite-volume formulation of the mass, momentum, energy and species conservation equations
and is implemented on block-structured grids.
Starting from an initial flow state with boundary conditions, state quantities in each finite-volume cell 
are updated in time to provide a simulation of the evolving flow field.
\end{abstract}

\clearpage

\tableofcontents

%------------------------------------------------------------------
\newpage
\section*{Nomenclature, Units}
%
\begin{tabular}{ll}
$A$	& : cell area, m$^2$ \\
$a$	& : speed of sound, m/s \\
$C_v$	& : specific heat at constant volume, $\rm J/kg \cdot K$ \\
$C_p$	& : specific heat at constant pressure, $\rm J/kg \cdot K$ \\
$E$	& : total energy per unit mass $e + \frac{1}{2} u^2$, J/kg \\
$e$	& : specific internal energy, J/kg \\
$\overline{F}$ & : array of flux terms \\
$f$            & : species mass fraction \\
$H$	& : total enthalpy, J/kg \\
$h$            & : specific enthalpy, J/kg \\
$\hat{i}, 
 \hat{j},
 \hat{k}$      & : unit vectors for the cartesian coordinates \\
$i,j,k$	& : cell index \\
$k$            & : coefficient of thermal conductivity \\
$M$	& : Mach number \\
$MW$	& : molecular weight \\
$n$            & : direction cosine \\
$\hat{n}, 
 \hat{p},
 \hat{q}$      & : unit vectors for the cell interface \\
$P, p$	& : pressure, Pa \\
$Pr$	& : Prandtl number \\
$Q$	& : source vector in the gas-dynamic equations \\
$q$     & : heat flux, W/m$^2$ \\
$R$	& : gas constant, $\rm J/kg \cdot K$ \\
$R_0$	& : universal gas constant, $\rm 8.314~J/mole \cdot K$ \\
$Re$	& : Reynolds number \\
$r$            & : radial coordinate, m \\
$S$            & : control surface of the cell \\
$T$	& : temperature, K \\
$t$	& : time, s \\
$U$	& : state vector in the gas-dynamic equations \\
$u$            & : velocity, m/s \\
$V$            & : cell volume, m$^3$ \\
$x, y, z$      & : cartesian coordinates, m \\
\end{tabular}

\begin{tabular}{ll}
$\alpha$	& : weighting parameter \\
$\Delta\pm$     & : intermediate variable for interpolation \\
$\lambda$ & : second coefficients of viscosity, Pa.s \\
$\mu$		& : viscosity, Pa.s; friction coefficient \\
$\rho$         & : density, kg/m$^3$ \\
$\gamma$       & : ratio of specific heats \\
$\tau$		& : wall shear stress, Pa \\
\end{tabular}


\subsection*{Subscripts}
\begin{tabular}{ll}
$i$     & : cell index, inviscid \\
$j,k$	& : cell indices \\
$L,~R$	& : left and right states for the Riemann solver or flux calculator \\
$max$	& : maximum value \\
$n$        & : normal to the cell interface \\
$p,q$      & : tangent to the cell interface \\
$s$	& ~~/ species index \\
$v$        & : viscous \\
$x,y,z$  & : coordinate directions \\
\end{tabular}


\subsection*{Superscripts}
\begin{tabular}{ll}
$\overline{(\cdot\cdot\cdot)}$ & : cell average \\
\end{tabular}


%------------------------------------------------------------------
\clearpage
\baselineskip = 1.5pc

\section{Introduction}
%
\texttt{Eilmer3} is a derivative of the code \texttt{mbcns2}, 
an acronym for Multiple-Block Compressible Navier-Stokes solver, second version.  
The code solves the compressible Navier--Stokes equations in order to provide simulations 
of transient compressible flow in two- and three-dimensions.
The governing equations are expressed in integral form over cell-centred, finite-volume cells, 
with the time rate of change of conserved quantities in each cell specified as a summation of 
the mass, momentum and energy flux through the cell interfaces.  

\medskip
The \texttt{mbcns2} code was an experiment in writing the \texttt{mb\_cns} code in C++.
\texttt{mb\_cns} (written in C) was derived from \texttt{cns4u}, 
a code written at ICASE in 1990-91 to simulate high-performance shock tunnels and expansion tubes.
The finite-difference codes of the 1980s did not do a good job of capturing the strong
shocks that processed the gas in these machines so we started work on a new code, 
using (what was at the time) the recently-proven upwinding approach.

\medskip
From the beginning, the code was intended to run on a parallel computer with 
blocks of finite-volume cells being allocated to the compute nodes of the Intel Hypercube computer
that had been acquired by ICASE in 1990.
However, in 1991, the code remained as a single-block code and was run on the 
Cray vector supercomputers at Langley to compute ideal gas flows in expansion tubes.
The code was used and further developed at UQ through the 1990s and into the 2000s.
Enhancements included:
\begin{itemize}
 \item more general thermochemistry, including look-up tables for gas in chemical equilibrium;
 \item multiple-block capability with an MPI parallel implementation;
 \item a parametric and programmable front-end for specifying geometry and grids;
 \item three-dimensional geometry; 
 \item programmable boundary conditions; and
 \item finite-rate chemistry.
\end{itemize}

\medskip
Once it was determined that there were clear benefits in using C++ \texttt{mb\_cns2},
our three-dimensional flow code Elmer was then reworked in C++ as \texttt{Elmer2}.
Of course, these codes being experiments in C++, we soon decided that it could all be done
much more cleanly be made much more versatile if we just reworked some of the basic modules.
Thus, the thermo-chemistry was reworked and the separate 2- and 3-dimensional codes merged
into Eilmer3.
The name change from \texttt{Elmer} to \texttt{Eilmer} is to a void a naming clash with the Elmer finite-element code 
from Finland.\footnote{http://www.csc.fi/elmer} 

\medskip
\texttt{Eilmer3} is actually an integrated collection of programs that includes a preparation program 
that can be used to set up a database of simulation parameters, 
a block-structured grid defining the flow domain and an initial flow field.
These items are then used as a starting point for the main simulation program
which computes a series of snapshots of the evolving flow.

\medskip 
The following sections describe the formulation of the code in terms of the basic gas dynamic model
and the thermochemical model for a multiple-species gas with finite-rate chemical kinetics.
There is a companion report \cite{jacobs_gollan_2010a} that desctibes the use of the code and provides 
a number of case studies.


%--------------------------------------------------------------------

% \clearpage
\input{gas-dynamics}

% \clearpage
\input{thermochem}

%--------------------------------------------------------------------
\clearpage
\bibliographystyle{unsrt}
\bibliography{../user-guide/bibtex/pj,../user-guide/bibtex/computing,../user-guide/bibtex/upwind,../user-guide/bibtex/gas_dynamic,../user-guide/bibtex/adm,../user-guide/bibtex/gas,../user-guide/bibtex/dan,../user-guide/bibtex/parallel,../user-guide/bibtex/rg-thesis,../user-guide/bibtex/shocktube}

\end{document}
