% elmer3-turbulence-model-validation.tex
% Overall LaTeX format adapted from PJ's elmer3 user guide
% 
% Wilson Chan, November 2008
%

\documentclass[12pt,a4paper]{article}
\usepackage[body={16cm,24.5cm}]{geometry}
\usepackage{pstricks}
\usepackage{graphicx}
\usepackage{listings}
\lstset{basicstyle=\scriptsize,identifierstyle=,keywordstyle=}

%------------------------------------------------------------------
% a couple horizontal bars to delimit embedded code
% the width suits the page size set above and
% the mathmode eliminates spaces between the three elements
\newcommand{\topbar}{\ensuremath{
    \rule{0.1mm}{2.0mm} \rule[2.0mm]{159.5mm}{0.1mm} \rule{0.1mm}{2.0mm}
}}
\newcommand{\bottombar}{\ensuremath{
    \rule{0.1mm}{2.0mm} \rule{159.5mm}{0.1mm} \rule{0.1mm}{2.0mm}
}}
\newcommand{\topbarshort}{\ensuremath{
    \rule{0.1mm}{2.0mm} \rule[2.0mm]{149.5mm}{0.1mm} \rule{0.1mm}{2.0mm}
}}
\newcommand{\bottombarshort}{\ensuremath{
    \rule{0.1mm}{2.0mm} \rule{149.5mm}{0.1mm} \rule{0.1mm}{2.0mm}
}}
%------------------------------------------------------------------

\title{
    The $k$-$\omega$ turbulence model in Eilmer3: \\
    User guide and test cases
}
\author{
    The University of Queensland \\
    School of Mechanical \& Mining Engineering \\
    Research Report Number 2010/01 
    \vspace{0.5cm} \\
    W. Y. K. Chan\thanks{Centre for Hypersonics, The University of 
     Queensland, Brisbane, Australia.}, 
    P. A. Jacobs\thanks{Queensland Geothermal Energy Centre of 
     Excellence, The University of Queensland, Brisbane, Australia.}, 
    J. P. Nap\thanks{Engineering Fluid Dynamics Group, University of Twente, Netherlands.}, \\ 
    D. J. Mee\thanks{School of Mechanical \& Mining Engineering, 
     The University of Queensland, Brisbane, Australia.},  
    R. M. Kirchhartz\thanks{Centre for Hypersonics, The University of
     Queensland, Brisbane, Australia.} and
    S. J. Stennett\thanks{Supervised by W. Y. K. Chan at the University of Queensland, Brisbane, Australia.} \\
}
% \date{November 2008}

\usepackage{subfigure}   % Adds capability to input sub-figures

\begin{document}
\maketitle

% For some reason, the abstract environment causes the labelling in
% the metapost figures to disappear; strange!
% \begin{abstract}
% \end{abstract}
\centerline{\textbf{Abstract}}
\medskip
This report provides details on the use and validation of Wilcox's
2006 $k$-$\omega$ turbulence model in Eilmer3. Eilmer3 is an integrated
collection of programs for the simulation of transient compressible flow
in two- and three-dimensions\cite{Jacobs2008,Jacobs2010}. The two-dimensional 
version of the $k$-$\omega$ model in Eilmer3 has been verified and
validated against four test cases. These cases include a two-dimensional 
flat plate case, an axisymmetric cylinder case, a backward-facing step
case and a coaxial-jets mixing case. The three-dimensional version of this turbulence model has also been investigated using three test cases, including a three-dimensional flat plate, single fin and normal injector. 
Since guidelines with regards to the setting up of grids in the
near-wall regions and specification of freestream turbulence properties 
for turbulent CFD simulations are rather scattered in literature, this 
report provides a compilation and, more importantly, a review of these 
recommended guidelines. 
\newpage
\tableofcontents

%-------------------------------------------------------------------
\include{./chap1-introduction/introduction}

% 2D examples
\include{./chap2-flat-plate/flat-plate}
\include{./chap3-mallinson-cylinder/mallinson-cylinder}
\include{./chap4-backward-facing-step/backward-facing-step}
\include{./chap5-coaxial-jets/coaxial-jets}

% 3D examples
\include{./chap6-3Dflatplate/3Dflatplate}
\include{./chap7-3Dsinglefin/3Dsinglefin}
\include{./chap8-3Dinjector/3Dinjector}

\include{./chap9-conclusion/conclusion}

%--------------------------------------------------------------------

\newpage
\bibliographystyle{unsrt}
%\bibliography{bibtex/pj,bibtex/computing,bibtex/gas_dynamic,bibtex/adm}
\bibliography{./bibliography/phd_bibliography}
%\bibliographystyle{newapa}

\newpage
\appendix

\end{document}
